PyPy: raising the level for Dynamic Language implementations

### 2007-05-27 talk/dyla2007/dyla.tex
Shrink shrink shrink the start of section 2.  It says less, but I think that the point that we want implementations on top of OO VMs was made in the introduction already.

The efforts required to build a new virtual machine are relatively
important.  This is particularly true for languages which are complex
and in constant evolution. Language implementation communities from an
open-source or academic context have only limited resources. Therefore they
cannot afford to have a highly complex implementation and often chose simpler
techniques even if that entails lower execution speed. Similarly fragmentation
(for example because of other implementations of the same language) is a
problem because it divides available resources. All these points also apply to
the implementation of domain-specific-languages where it is important to keep
implementation effort small.

@@

XXX the next paragraph is also the start of the introduction right now!

Dynamic languages are traditionally implemented by writing a virtual
machine for them in a low-level language like C, or in a language that
can relatively easily be turned into C.  The machine implements an
object model supporting the high level dynamic language's objects.  It
typically provides features like automatic garbage collection.  Recent
languages like Python, Ruby, Perl and JavaScript have complicated
semantics which are most easily mapped to a simple interpreter operating
on syntax trees or bytecode. Simpler languages like Lisp and Self
typically have more efficient implementations based on just-in-time code
generation.

There are many good reasons for choosing C as an implementation language. It
gives a decent base performance, is highly portable across different hardware
architectures and operating system if some care is taken and it allows the reuse
of many existing libraries.

On the other hand the use of C comes with a set of problems. Since C has no ways
to do automatic memory management and since the implemented languages are
garbage collected, the VM needs to contain a garbage collector. Implementing a
high-performance GC is a tedious task and so many implementations resort to
relatively simple algorithms like mark-and-sweep or reference counting. Also the
choice of garbage collector is fixed relatively early in the life of the project
and then manifests itself throughout the complete VM source code. Thus it
becomes extremely tedious to change it later.

%A similar problem is the threading model the VM uses
%  - simpler threading models
%    - don't make single-threaded apps slower due to locks
%    - global locks
%    - green threading

% - wrapping with C libraries simple but tedious

These limitations of a C-based implementation often lead to more implementations
of the language. For example in addition to the main C implementation of Python
(``CPython'' \cite{cpy251}) there are a number of additional implementations
that want to fix some of the limitations of CPython. One of them is ``Stackless
Python'' \cite{stackless} which is a fork of CPython that adds micro-threading
capabilities to Python. One of the reasons for not incorporating it back into
CPython was that it was felt that they would make the implementation too
complex. Another implementation of the Python language is the Psyco
\cite{psyco-software}
project which adds a JIT-compiler to CPython. As all other re-implementations,
Psyco contains the Python semantics and needs to be kept synchronized with
CPython manually.

% - free threading attempts for Python

As a language becomes more popular, yet another reason for different
implementations emerges: People want to use it together with Java and C\#
libraries and tools. Although it is in principle possible to do that by
bridging the C-VM with the respective VMs of these languages (for example the
``PythonNet'' project \cite{pythonnet} that brigdes CPython with a .NET VM)
this is an incomplete solution since it prevents many use-cases.
Therefore in the end implementations of the dynamic language on top of the
object-oriented VMs are created to properly integrate with the complete
environment.

@@

Implementing a language on top of an OO VM is in many ways easier than
implementing it in C. Let's take a look at the advantages that are usually
cited for basing a
language implementation of a dynamic language on a standard object oriented
virtual machine, for example the JVM or the CLR. The central theme of the
benefits  of OO VMs is the ability to implement certain hard things only once
and share the benefits between all language implementations on top of the OO VM.

@@

\begin{itemize}
\item
\emph{Better performance:} So far it seems that performance of highly dynamic
languages is not actually significantly improved on OO VMs. 
Jython\footnote{Python on the Java VM} is around 5
times slower than CPython, for IronPython\footnote{Python on .NET, which
gives up on some features to improve performance}
the figures vary but it is mostly
within the same order of magnitude as CPython. The most important reason for
this is that the VM's JIT compilers are optimized for specific usage patterns
that are common in the main language of the OO VM. To get good speeds the
language implementers would have to carefully produce code that matches these
usage patterns, which is not a simple task.

>>>>

The efforts required to build a new virtual machine are relatively
important.  This is particularly true for languages which are complex
and in constant evolution. Language implementation communities from an
open-source or academic context have only limited resources. Therefore they
cannot afford to have a highly complex implementation and often chose simpler
techniques even if that entails lower execution speed. Similarly fragmentation
(for example because of other implementations of the same language) is a
problem because it divides available resources. All these points also apply to
the implementation of domain-specific languages where it is important to keep
the implementation effort small.

@@

\def\implname#1{\emph{#1}}

The fact that limitations of a C-based implementation of a dynamic
language leads to the emergence of additional implementations is clear
in the case of Python.  The reference implementation, \implname{CPython}
\cite{cpy251}, is a simple recursive interpreter.  \implname{Stackless
Python} \cite{stackless} is a fork that adds micro-threading
capabilities to Python. One of the reasons for not incorporating it back
into CPython was that it was felt that they would make the
implementation too complex. Another implementation of the Python
language is \implname{Psyco} \cite{psyco-software}, which adds a
JIT-compiler to CPython.  Finally, \implname{Jython} is a
re-implementation for the Java VM and \implname{IronPython} one for
.NET.  All of these need to be kept in sync with the relatively fast
evolution of the language.

With the emergence of .NET and the JVM as interesting language
implementation platforms, an argument that is sometimes made is that
communities should only develop an implementation of their language for
one of these platforms.\footnote{Preferably the argument author's favourite
one.}

@@

Implementing a language on top of an existing OO VM is in many ways easier than
implementing it in C. Let's take a look at the advantages that are usually
cited for basing a
language implementation of a dynamic language on a standard object oriented
virtual machine, for example the JVM or the CLR. The central theme of the
benefits  of OO VMs is the ability to implement certain hard things only once
and share the benefits between all language implementations on top of the OO VM.

@@

\begin{itemize}
\item
\emph{Better performance:} So far it seems that performance of highly dynamic
languages is not actually significantly improved on OO VMs. 
Jython is around 5
times slower than CPython, for IronPython\footnote{Python on .NET, which
gives up on some features to improve performance}
the figures vary but it is mostly
within the same order of magnitude as CPython. The most important reason for
this is that the VM's JIT compilers are optimized for specific usage patterns
that are common in the main language of the OO VM. To get good speeds the
language implementers would have to carefully produce code that matches these
usage patterns, which is not a simple task.
